### 2022-02-23 paper/localpsf.tex
updating paper based on advances the last couple months. Work in progress...

Let $\Omega \subset \mathbb{R}^d$ be a bounded domain (in applications typically $\gdim=1$, $2$, or $3$). We consider integral operators $\Aop:L^2(\Omega)\rightarrow L^2(\Omega)'$ of the form
\begin{equation}
	\label{eq:kernel_representation}
	(\Aop u)(v) := \int_\Omega \int_\Omega v(y) \Aker(y,x) u(x) dx dy,
\end{equation}
where $\Aker:\Omega \times \Omega \rightarrow \mathbb{R}$ is the integral kernel. Here $\Aop u \in L^2(\Omega)'$ is the linear functional that results from applying $\Aop$ to $u\in L^2(\Omega)$, and $\left(\Aop u\right)(v)$ is the scalar that results from applying that linear functional to $v \in L^2(\Omega)$. Further, let $\impulseresponse_x:\Omega \rightarrow \mathbb{R}$,
\begin{equation}
\label{eq:impulse_response_defn}
\impulseresponse_x(y) := \Aker(y, x),
\end{equation} 
be the \emph{impulse response} of $\Aop$ at $x$. The function $\impulseresponse_x$ is called the impulse response because straightforward analysis shows that it may be expressed as follows:
\begin{equation}
\label{eq:impulse_response_delta_action}
\impulseresponse_x = \left( \Aop \delta_x^* \right)^*,
\end{equation}
where $\delta_x$ is the the point source (delta distribution) centered at $x$, $\Aop \delta_x^* \in L^2(\Omega)'$ is an abuse of notation that denotes the result of applying the operator $\Aop$ to $\delta_x$, and $\left( \Aop \delta_x^* \right)^* \in L^2(\Omega)$ denotes the Riesz representation of $\Aop \delta_x^*$ with respect to the $L^2$ inner product.\footnote{Recall that the Riesz representative of a functional $\sigma \in L^2(\Omega)'$ with respect to the $L^2$ inner product is the unique function $\sigma^* \in L^2(\Omega)$ such that $\sigma(v) = \left(\sigma^*,v\right)_{L^2(\Omega)}$ for all $v \in L^2(\Omega)$.} Note that while the domain of $\Aop$ is formally defined as $L^2(\Omega)$, which does not contain $\delta_x$, if $\Aker$ is sufficiently regular then the action of $\Aop$ may be extended to distributions. This is described in Appendix \ref{app:distributions}.

We seek to approximate operators $\Aop$ with locally supported, non-negative impulse responses that are accessible matrix-free. 
\begin{itemize}
	\item $\Aop$ has \textbf{locally supported impulse responses} if, for all $x\in \Omega$, the support of $\impulseresponse_x$ is contained (or approximately contained) in a small region.
	\item  $\Aop$ has \textbf{non-negative impulse responses} if
	\begin{equation*}
	\impulseresponse_x(y) \ge 0
	\end{equation*}
	for all (or approximately all) $(y,x) \in \Omega \times \Omega$.
	\item $\Aop$ is accessible \textbf{matrix-free} if one cannot easily evaluate kernel entries $\Aker(y,x)$. Instead, access to $\Aop$ is available through ``operator actions.'' That is, we have a black-box computational procedure (which is typically expensive) through which we may evaluate the maps
	\begin{equation*}
	u \mapsto\Aop u \quad \text{and} \quad v \mapsto \Aop^Tv
	\end{equation*} 
	for arbitrary functions $u$ and $v$. After discretization (see Appendix \ref{app:discretized_operations}), $\Aop$ becomes a dense matrix, $\bm{A}$, that is typically too large to be built and stored. This property is called ``matrix-free'' because at the discrete level it means that one can perform matrix-vector products of $\bm{A}$ and $\bm{A}^T$ with arbitrary vectors, but one cannot easily\footnote{One could compute $\bm{v} = \bm{A}\bm{e}_j$, where $\bm{e}_j=(0,\dots,0,1,0,\dots,0)$ is the unit vector which has $j^\text{th}$ component equal to one and all other components equal to zero, then extract $\bm{A}_{ij} = \bm{v}_i$. However, this process is wasteful because one computes the entire vector $\bm{v}$, then discards all but one component.} access matrix entries $\bm{A}_{ij}$. In applications, applying $\bm{A}$ to a vector may involve iteratively solving a large linear system, timestepping, or performing some other nontrivial computational procedure. 
	%For example, in reduced space approaches to PDE-constrained optimization and inverse problems, Hessian information is available only through the application of the Hessian to vectors, and one such Hessian application requires solving two PDEs \cite{HESSIANADJOINT}. 
\end{itemize}

@@

\subsection{Idea of the method}

Because of the non-negative impulse response property, $\impulseresponse_x$ may be interpreted as a scaled probability distribution. Let $\spatialvol(x) := \int_\Omega \impulseresponse_x(y) dy$ denote the scaling factor (the ``volume'' of $\impulseresponse_x$), let $\widehat{\impulseresponse}_x := \impulseresponse_x / \spatialvol(x)$ denote the normalized version of $\impulseresponse_x$, and let $\spatialmean(x)$ and $\spatialcov(x)$ denote the mean and covariance of $\widehat{\impulseresponse}_x$, respectively. In Section \ref{eq:mean_and_covariance_estimation} we will see that it is possible to compute $V(x)$, $\mu(x)$, and $\Sigma(x)$ \emph{for all $x \in \Omega$ simultaneously}, via a procedure that involves applying $\Aop^T$ to a small number of specially chosen functions (three functions if $\gdim=1$, six if $\gdim=2$, ten if $\gdim=3$). Knowing $V(x)$, $\mu(x)$, and $\Sigma(x)$ allows to form an ellipsoid shaped estimate for the support of $\phi_x$. We can predict where $\phi_x$ will be nonzero \emph{before} we compute it.

Our overall strategy is to compute impulse responses $\{\impulseresponse_{x_i}\}_{i=1}^r$ of $\Aop$ to point sources at a collection of points $x_i$ using a small number of operator actions of $\Aop$. If $\Aop$ is non-symmetric, we also compute impulse responses $\{\impulseresponse'_{y_i}\}_{i=1}^r$ of $\Aop^T$ to point sources at a collection of points $y_i$. We then form a H-matrix approximation, $\boldsymbol{\Phi}_H$, of the kernel, $\Phi$, and use H-matrix methods to convert $\boldsymbol{\Phi}_H$ into an H-matrix approximation of $\Aop$, denoted $\mathbf{A}_H$. Forming the H-matrix approximation $\boldsymbol{\Phi}_H$ requires evaluating kernel entries, $\Phi(y,x)$, at arbitrary points $(y,x)$, and this is done by interpolating the known impulse responses $\{\impulseresponse_{x_i}\}_{i=1}^r, \{\impulseresponse'_{y_i}\}_{i=1}^r$. The details of the interpolation will be described in detail later, but underlying idea behind the interpolation is that 
$$\Phi(y,x) = \phi_x(y) \approx \phi_{x_i}(y - \mu(x) + \mu(x_i))$$
if $x_i$ is close to $x$.
This is shown in Figure FIG.
Schematically the overall framework proceeds as follows:
\begin{equation*}
	\Aop \longrightarrow V, \mu, \Sigma \longrightarrow \{\impulseresponse_{x_i}\}_{i=1}^r, \{\impulseresponse'_{y_i}\}_{i=1}^r \longrightarrow \boldsymbol{\Phi}_H \longrightarrow \mathbf{A}_H.
\end{equation*}
We will define product-convolution approximations in Section \ref{sec:prodconv_intro}. H-matrices are discussed in Appendix \ref{app:h_matrix}. Our method is summarized in Section \ref{sec:overview_intro}, and described in detail in Section \ref{sec:method}.

The novel contributions of our method are in the way that we form the product-convolution approximation ($\Aop \rightarrow \AopPc$). This is the part of the method where we use the four properties of $\Aop$ described above. Fast conversion of the discretized version of $\AopPc$ to H-matrix format ($\AopPc \rightarrow \AmatPc$) is possible because integral kernel entries for $\AopPc$ are easily accessible (unlike integral kernel entries for $\Aop$). Once in H-matrix format, one may use H-matrix methods to perform further useful linear algebra operations involving $\AmatPc$, including matrix-matrix addition, matrix-matrix multiplication, matrix-vector products, matrix factorization, and matrix inversion. These H-matrix methods are fast and scale well to large problems.

@@

>>>>

Let $\Omega \subset \mathbb{R}^d$ be a bounded domain (in applications typically $\gdim=1$, $2$, or $3$). We consider integral operators $\Aop:L^2(\Omega)\rightarrow L^2(\Omega)'$ of the form
\begin{equation}
	\label{eq:kernel_representation}
	(\Aop u)(v) := \int_\Omega \int_\Omega v(y) \Aker(y,x) u(x) dx dy,
\end{equation}
where $\Aker:\Omega \times \Omega \rightarrow \mathbb{R}$ is the integral kernel. Here $\Aop u \in L^2(\Omega)'$ is the linear functional that results from applying $\Aop$ to $u\in L^2(\Omega)$, and $\left(\Aop u\right)(v)$ is the scalar that results from applying that linear functional to $v \in L^2(\Omega)$.

Further, let $\impulseresponse_x:\Omega \rightarrow \mathbb{R}$,
\begin{equation}
\label{eq:impulse_response_defn}
\impulseresponse_x(y) := \Aker(y, x),
\end{equation} 
be the \emph{impulse response} of $\Aop$ at $x$. The function $\impulseresponse_x$ is called the impulse response because straightforward analysis shows that it may be expressed as follows:
\begin{equation}
\label{eq:impulse_response_delta_action}
\impulseresponse_x = \left( \Aop \delta_x^* \right)^*,
\end{equation}
where $\delta_x$ is the the point source (delta distribution) centered at $x$, $\Aop \delta_x^* \in L^2(\Omega)'$ is an abuse of notation that denotes the result of applying the operator $\Aop$ to $\delta_x$, and $\left( \Aop \delta_x^* \right)^* \in L^2(\Omega)$ denotes the Riesz representation of $\Aop \delta_x^*$ with respect to the $L^2$ inner product.\footnote{Recall that the Riesz representative of a functional $\sigma \in L^2(\Omega)'$ with respect to the $L^2$ inner product is the unique function $\sigma^* \in L^2(\Omega)$ such that $\sigma(v) = \left(\sigma^*,v\right)_{L^2(\Omega)}$ for all $v \in L^2(\Omega)$.} Note that while the domain of $\Aop$ is formally defined as $L^2(\Omega)$, which does not contain $\delta_x$, if $\Aker$ is sufficiently regular then the action of $\Aop$ may be extended to distributions. This is described in Appendix \ref{app:distributions}.

In this paper we form hierarchical matrix approximations of operators $\Aop$ with locally supported, non-negative impulse responses that are accessible matrix-free.
\begin{description}
	\item[Locally supported impulse responses:] For all $x\in \Omega$, the support of $\impulseresponse_x$ is contained (or approximately contained) in a localized region.
	\item[Non-negative impulse responses:] We have
	\begin{equation*}
	\impulseresponse_x(y) \ge 0
	\end{equation*}
	for all (or approximately all) $(y,x) \in \Omega \times \Omega$.
	\item[Matrix-free:] One cannot easily evaluate kernel entries $\Aker(y,x)$. Instead, access to $\Aop$ is available through ``operator actions.'' That is, we have a black-box computational procedure (which is typically expensive) through which we may evaluate the maps
	\begin{equation*}
	u \mapsto\Aop u \quad \text{and} \quad v \mapsto \Aop^Tv
	\end{equation*} 
	for arbitrary functions $u$ and $v$. After discretization (see Appendix \ref{app:discretized_operations}), $\Aop$ becomes a dense matrix, $\bm{A}$, that is typically too large to be built and stored. This property is called ``matrix-free'' because at the discrete level it means that one can perform matrix-vector products of $\bm{A}$ and $\bm{A}^T$ with arbitrary vectors, but one cannot easily\footnote{One could compute $\bm{v} = \bm{A}\bm{e}_j$, where $\bm{e}_j=(0,\dots,0,1,0,\dots,0)$ is the unit vector which has $j^\text{th}$ component equal to one and all other components equal to zero, then extract $\bm{A}_{ij} = \bm{v}_i$. However, this process is wasteful because one computes the entire vector $\bm{v}$, then discards all but one component.} access matrix entries $\bm{A}_{ij}$. In applications, applying $\bm{A}$ to a vector may involve iteratively solving a large linear system, timestepping, or performing some other nontrivial computational procedure. 
	%For example, in reduced space approaches to PDE-constrained optimization and inverse problems, Hessian information is available only through the application of the Hessian to vectors, and one such Hessian application requires solving two PDEs \cite{HESSIANADJOINT}. 
\end{description}

@@

@@

\subsection{Idea of the method}
Each impulse response $\impulseresponse_x$ may be interpreted as a scaled probability distribution because of the non-negative impulse response property. Let $\spatialvol(x) := \int_\Omega \impulseresponse_x(y) dy$ denote the scaling factor (the ``volume'' of $\impulseresponse_x$), let $\widehat{\impulseresponse}_x := \impulseresponse_x / \spatialvol(x)$ denote the normalized version of $\impulseresponse_x$, and let $\spatialmean(x)$ and $\spatialcov(x)$ denote the mean and covariance of $\widehat{\impulseresponse}_x$, respectively. In Section \ref{eq:mean_and_covariance_estimation} we will see that it is possible to compute $V(x)$, $\mu(x)$, and $\Sigma(x)$ for all $x \in \Omega$ simultaneously, via a procedure that involves applying $\Aop^T$ to a small number of polynomial functions (three functions if $\gdim=1$, six if $\gdim=2$, ten if $\gdim=3$). Since a probability distribution concentrates within a small number of standard deviations of its mean, knowing $\mu(x)$ and $\Sigma(x)$ permits us to form an ellipsoid shaped a-priori estimate for the support of $\phi_x$. In other words, we can predict an ellipsoid outside of which $\phi_x$ will be (nearly) zero, \emph{without} computing $\phi_x$.

This allows us to compute large numbers of impulse responses, $\impulseresponse_{x_i}$ in ``batches.'' We compute one batch of $\impulseresponse_{x_i}$'s by applying $\Aop$ to a weighted sum of point sources (Dirac comb) associated with a collection of points $x_i$ scattered throughout $\Omega$. The batch of points $x_i$ are chosen via a greedy ellipsoid packing algorithm so that the support of $\impulseresponse_{x_i}$ and the support of $\impulseresponse_{x_j}$ do not overlap (or do not overlap much) if $i \neq j$. Because these supports do not overlap, we can post-process the response of $\Aop$ to the Dirac comb to recover the functions $\impulseresponse_{x_i}$ associated with all points $x_i$ in the batch---with one application of $\Aop$, we recover many $\impulseresponse_{x_i}$. This is illustrated in Figure FIG. The process is repeated to get more batches, until a desired number of impulse responses is reached.

Once the impulse responses $\{\impulseresponse_{x_i}\}_{i=1}^r$ are computed, we may approximate the integral kernel $\Phi(y,x)$ at arbitrary points $(y,x)$ by interpolation. The details of the interpolation will be described in detail later, but the key idea is that
$$\Phi(y,x) = \phi_x(y) \approx \phi_{x_i}(y - \mu(x) + \mu(x_i))$$
if $x_i$ is close to $x$. This is shown in Figure FIG. We therefore approximate $\phi_x(y)$ by a weighted linear combination of the values $\phi_{x_i}(y - \mu(x) + \mu(x_i))$ for the $k$-nearest points $x_i$ to $x$. Here $k$ is a user-defined parameter; we use $k=10$. The weights in the linear combination are determined via radial basis function interpolation.

The ability to rapidly evaluate approximations to arbitrary kernel entries $\Phi(y,x)$ allows us to construct an H-matrix representation of $\Phi$, which we denote $\boldsymbol{\Phi}_H$, using the conventional adaptive cross H-matrix construction method. In this method, one forms low-rank approximations of blocks of the matrix by sampling rows and columns of those blocks \cite{HACA}. We then use H-matrix methods to convert $\boldsymbol{\Phi}_H$ into an H-matrix approximation of $\Aop$, denoted $\mathbf{A}_H$. Schematically the overall framework proceeds as follows:
\begin{equation*}
\Aop \longrightarrow V, \mu, \Sigma \longrightarrow \{\impulseresponse_{x_i}\}_{i=1}^r \longrightarrow \boldsymbol{\Phi}_H \longrightarrow \mathbf{A}_H.
\end{equation*}
Once in H-matrix format, one may use H-matrix methods to perform further useful linear algebra operations involving $\mathbf{A}_H$, including matrix-matrix addition, matrix-matrix multiplication, matrix-vector products, matrix factorization, and matrix inversion. These H-matrix methods are fast and scale well to large problems.

Often we are interested in approximating operators $\Aop$ that are symmetric positive semi-definite. Unfortunately, the approximation is, in general, non-symmetric and indefinite due to errors in the approximation. In this case, we modify the H-matrix representation of $\mathbf{A}_H$ to make it symmetric positive definite, by symmetrizing it, then applying a specially chosen rational matrix function to the symmetrized H-matrix.

%Forming the H-matrix approximation $\boldsymbol{\Phi}_H$ requires evaluating kernel entries, $\Phi(y,x)$, at arbitrary points $(y,x)$, and this is done by interpolating the known impulse responses $\{\impulseresponse_{x_i}\}_{i=1}^r, \{\impulseresponse'_{y_i}\}_{i=1}^r$. 
%
%%For each batch, we solve an ellipsoid packing problem to choose as many points $x_i$ as possible, while ensuring that the supports of the impulse responses $\impulseresponse_{x_i}$ do not overlap for points $x_i$ in the batch. Then we apply $\Aop$ to a weighted sum of delta distributions centered at the points $x_i$ (Dirac comb). From the resulting 
%
%
%Our overall strategy is as follows. First, we compute the functions $V:\Omega \rightarrow \mathbb{R}$, $\mu:\Omega \rightarrow \mathbb{R}^d$, $\Sigma:\Omega \rightarrow \mathbb{R}^{d \times d}$ by applying $\Aop^T$ to a small number of polynomial functions. Next, we solve ellipsoid packing problems to choose batches of points $x_i$ such that 
%to compute impulse responses $\{\impulseresponse_{x_i}\}_{i=1}^r$ of $\Aop$ to point sources at a collection of points $x_i$ using a small number of operator actions of $\Aop$. If $\Aop$ is non-symmetric, we also compute impulse responses $\{\impulseresponse'_{y_i}\}_{i=1}^r$ of $\Aop^T$ to point sources at a collection of points $y_i$. We then form a H-matrix approximation, $\boldsymbol{\Phi}_H$, of the kernel, $\Phi$, and use H-matrix methods to convert $\boldsymbol{\Phi}_H$ into an H-matrix approximation of $\Aop$, denoted $\mathbf{A}_H$. 
%Schematically the overall framework proceeds as follows:
%\begin{equation*}
%\Aop \longrightarrow V, \mu, \Sigma \longrightarrow \{\impulseresponse_{x_i}\}_{i=1}^r \longrightarrow \boldsymbol{\Phi}_H \longrightarrow \mathbf{A}_H.
%\end{equation*}
%We will define product-convolution approximations in Section \ref{sec:prodconv_intro}. H-matrices are discussed in Appendix \ref{app:h_matrix}. Our method is summarized in Section \ref{sec:overview_intro}, and described in detail in Section \ref{sec:method}. 
%
%The novel contributions of our method are in the way that we form the product-convolution approximation ($\Aop \rightarrow \AopPc$). This is the part of the method where we use the four properties of $\Aop$ described above. Fast conversion of the discretized version of $\AopPc$ to H-matrix format ($\AopPc \rightarrow \AmatPc$) is possible because integral kernel entries for $\AopPc$ are easily accessible (unlike integral kernel entries for $\Aop$). Once in H-matrix format, one may use H-matrix methods to perform further useful linear algebra operations involving $\AmatPc$, including matrix-matrix addition, matrix-matrix multiplication, matrix-vector products, matrix factorization, and matrix inversion. These H-matrix methods are fast and scale well to large problems.
